---
title: "Grass/PDO model selection"
author: "Erica Christensen"
date: "8/11/2021"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(MuMIn)

# min-max normalization function
normalize <- function(x) {
  return ((x-min(x)) / (max(x) - min(x)))
}

# 10 year moving average function
moving_avg_10y <- function(index, year) {
  smoothed = c()
  for (n in 10:length(index)) {
    avg10y = data.frame(index_smooth = mean(index[(n-9):n]),
                        year=year[n])
    smoothed = rbind(smoothed, avg10y)
  }
  return(smoothed)
}


# read in data
#grass = read.csv('data/grass_shrub_trends_yearly.csv')
grass = read.csv('data/grass_shrub_median_yearly.csv') %>% rename(grass = median_grass)
pdoindex = read.csv('data/PDO_long_1900_2020.csv')
soiindex = read.csv('data/SOI_long_1866_2020.csv')
pptdat = read.csv('data/climate_variables.csv')
pdo_factor = read.csv('data/PDO_phases_byyear.csv')

# convert monthly indices to yearly
pdo_yearly = pdoindex %>% group_by(year) %>%
  summarize(pdo_year = mean(pdo))
soi_yearly = soiindex %>% group_by(year) %>%
  summarise(soi_year=mean(soi))


# normalize timeseries
pdo_yearly$pdo_norm = normalize(pdo_yearly$pdo_year)
soi_yearly$soi_norm = normalize(soi_yearly$soi_year)
grass$grass_norm = normalize(grass$grass)
pptdat$yearly_ppt_norm = normalize(pptdat$yearly_ppt_mm)
pptdat$winter_ppt_norm = normalize(pptdat$winter_ppt_mm)
pptdat$summer_ppt_norm = normalize(pptdat$summer_ppt_mm)

# calculate 10-year moving average of pdo
pdo_10y = moving_avg_10y(pdo_yearly$pdo_year, year=pdo_yearly$year) %>%
  mutate(pdo_smooth=normalize(index_smooth))

# calculate moving average of yearly precip
precip_10y = moving_avg_10y(pptdat$yearly_ppt_mm, year=pptdat$year) %>%
  mutate(precip_smooth=normalize(index_smooth))

# moving avg summer and winter precip
winter_ppt_10y = moving_avg_10y(pptdat$winter_ppt_mm, year=pptdat$year) %>%
  mutate(winter_ppt_smooth=normalize(index_smooth))
summer_ppt_10y = moving_avg_10y(pptdat$summer_ppt_mm, year=pptdat$year) %>%
  mutate(summer_ppt_smooth=normalize(index_smooth))

# create one big data frame
combined_data = merge(pdo_yearly, soi_yearly, all=T) %>%
  merge(grass, by.x='year', by.y='project_year', all=T) %>%
  merge(pptdat, all=T) %>% 
  merge(pdo_10y[,c('year','pdo_smooth')], all=T) %>%
  merge(pdo_factor, all=T) %>%
  merge(precip_10y[,c('year','precip_smooth')], all=T) %>%
  merge(winter_ppt_10y[,c('year','winter_ppt_smooth')], all=T) %>%
  merge(summer_ppt_10y[,c('year','summer_ppt_smooth')], all=T) %>%
  dplyr::select(year, grass_norm, pdo_norm, pdo_smooth, pdo_phase, soi_norm, yearly_ppt_norm, precip_smooth, winter_ppt_norm, winter_ppt_smooth,
                summer_ppt_norm, summer_ppt_smooth)
```

## Timeseries cross-correlation

The first thing I did was look at the ccf between grass (normalized) and the normalized climate variable time series. PDO had significant correlation terms going back to lag 10, while SOI, yearly precip, winter precip, and summer precip had no significant terms.

```{r cars, echo=F}
# cross-correlation functions
corr_data = dplyr::filter(combined_data, !is.na(grass_norm))
ccf(corr_data$pdo_norm, corr_data$grass_norm)
ccf(corr_data$soi_norm, corr_data$grass_norm)
ccf(corr_data$yearly_ppt_norm, corr_data$grass_norm)
ccf(corr_data$winter_ppt_norm, corr_data$grass_norm)
ccf(corr_data$summer_ppt_norm, corr_data$grass_norm)
```

## Models and model comparison

I created a smoothed PDO variable by calculating 10-year moving window average of the PDO index. I also created a variable indicating PDO phase (cool or warm; factor variable). I compared models correlating normalized grass cover to PDO index, the binary warm/cool pdo phase, 10-year smoothed PDO, SOI, yearly precip, winter precip, and summer precip.

```{r pressure, echo=T}
model_data = combined_data %>%
  dplyr::filter(year>=1916, year<1980)

# models with single predictors
model1 = lm(grass_norm ~ pdo_norm, data=model_data)
model2 = lm(grass_norm ~ pdo_phase, data=model_data)
model3 = lm(grass_norm ~ pdo_smooth, data=model_data)
model4 = lm(grass_norm ~ soi_norm, data=model_data)
model5 = lm(grass_norm ~ yearly_ppt_norm, data=model_data)
model6 = lm(grass_norm ~ winter_ppt_norm, data=model_data)
model7 = lm(grass_norm ~ summer_ppt_norm, data=model_data)

AICc(model1, model2, model3, model4, model5, model6, model7)
```

```{r multimodels, echo=F}
# models with multiple predictors

model8 = lm(grass_norm ~ pdo_smooth + yearly_ppt_norm, data=model_data)
model9 = lm(grass_norm ~ pdo_smooth + winter_ppt_norm, data=model_data)
model10 = lm(grass_norm ~ pdo_smooth + summer_ppt_norm, data=model_data)
model11 = lm(grass_norm ~ pdo_smooth + soi_norm, data=model_data)
model12 = lm(grass_norm ~ pdo_smooth + soi_norm + summer_ppt_norm, data=model_data)
model13 = lm(grass_norm ~ pdo_smooth + soi_norm + yearly_ppt_norm, data=model_data)
model14 = lm(grass_norm ~ pdo_smooth + soi_norm + winter_ppt_norm, data=model_data)

AICc(model3, model8, model9, model10, model11, model12, model13, model14)
```

The smoothed PDO (model3) is clearly the best. I then tested models containing smoothed PDO + another variable, but no model was more than 2 units better than model2

## Predictions from model

```{r model predict}
# prediction based on best model
newdata = dplyr::filter(combined_data, year>1916) %>%
  dplyr::select(year, pdo_smooth)
predicted_grass = predict.lm(model3, newdata, se.fit=T)

modelprediction = newdata %>%
  mutate(predicted_grass=predicted_grass$fit)

se = 0.08134 # from summary(model3)
# plot figure
grasspred = ggplot(modelprediction, aes(x=year)) +
  geom_line(aes(y=predicted_grass)) +
  geom_ribbon(aes(ymin=predicted_grass-se, ymax=predicted_grass+se), alpha=.3) +
  geom_point(data=combined_data, aes(x=year, y=grass_norm)) +
  xlim(1916, 2020) +
  theme_bw() +
  xlab('') +
  ylab('grass cover')
grasspred
ggsave('Figures/grass_prediction_from_PDO.png', plot=grasspred, width=5, height=4)
```